The `forestecology` R package for modeling interspecies competition between trees

### 2021-02-16 etc/paper/forestecology.Rmd
Done "Identify all focal and corresponding competitor trees" section

```{r setup, include = FALSE, cache = FALSE}
# Internally used packages
library(viridis)
library(knitr)
library(ggplot2)
library(here)

# knitr settings
opts_chunk$set(
  # Code output:
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  cache = TRUE,
  # Figure:
  out.width = "100%",
  fig.path = "Figures/",
  fig.width = 16 / 2.5,
  fig.height = 9 / 2.5,
  fig.align = "center",
  fig.show = "hold",
  # Etc:
  collapse = TRUE,
  comment = "##"
  # tidy = FALSE
)

@@

@@

1. Compute the growth of trees based on census data
1. Add spatial information:
    1. Define buffer region trees.
    1. Add spatial cross-validation block information.
1. Compute focal versus competitor tree information.
1. Fit model and make predictions.
1. Visualize model results.
1. Additionally: Evaluate model performance using spatial cross-validation.
1. Additionally: Evaluate the effect of competitor species identity using permutation tests.

@@

The first step in the our analysis sequence is to compute the growth of trees using data from two censuses. The `compute_growth()` function computes growth assuming census data that follows ForestGEO standards. Despite such standards, minor variations will still exist between sites thereby necessitating some data wrangling and checking. For example, the SCBI site records all dbh's in millimeters, whereas the Michigan Big Woods site records them in centimeters @andersonteixeira_ctfs-forestgeo_2015 @allen_michigan_2020. The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all trees is saved in a data frame of class `sf` whereby the geolocation of each tree is presented in a `geometry` variable of type `<POINT>` and at a minimum the data contains the following variables: a variable uniquely identifying each tree-stem, `sp` of type `fct` factor identifying species, `dbh1` and `dbh2` of type `dbl` quantifying the dbh at earlier and later census, and `growth` of type `dbl` double quantifying the average annual growth in centimeters.

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020. After selecting only relevant variables, we perform a few additional data wrangling steps: convert the variable with the date of measurement to be of type `date`, convert dbh to be in centimeters^[A rule of thumb to determine the units of dbh is check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with dbh greater or equal to 1cm should be included in censuses.], convert the `sp` variable containing species information from type `chr` character to `fct` factor (we will discuss the need for this in Section \ref{spatial-cross-validation}). Furthermore, in order to speed up computation for purposes of this examplt, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

The output `growth_scbi` is a single data frame of class `sf` that includes a numerical `growth` reflecting the average annual growth in dbh (in cm) for all trees that were alive at both time points as well a `geometry` variable encoding each tree's geolocation. Furthermore, variables that (in theory) remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`.

@@

## Define focal versus competitor trees

@@

Next we define `focal_vs_comp` data frames which connects each focal tree in the `growth_df` data frames to the trees in its competitive neighborhood range as defined by the `comp_dist` constant. So for example, if `growth_df` consisted of two focal trees with two and three neighbors with `comp_dist` respectively, `focal_vs_comp` would be a data frame of 5 rows connecting each focal tree to it's competitors. The `create_focal_vs_comp()` function makes this connection taking as inputs the `growth_df` data frame; the `comp_dist` constant defining competitive range; `cv_grid_sf`, giving the cross validation grid; and the `id` variable.

```{r bw-focal-vs-comp}
focal_vs_comp_bw <- growth_bw %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = blocks_bw, id = "treeID")
```

TODO: Figure out how to show this data frame's contents.

```{r scbi-focal-vs-comp}
focal_vs_comp_scbi <- growth_scbi %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = blocks_scbi, id = "stemID")
```

TODO: Figure out how to show this data frame's contents.

Next we fit the following linear model to the dbh of each focal tree. Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ "competitor" species groups. We modeled the growth in diameter per year $y_{ij}$ (in centimeters per year) of the $i^{th}$ tree of focal species group $j$ as a linear model $f$ of the following covariates $\vec{x}_{ij}$

@@

Using these posterior means, we then use the posterior predictive distribution to obtain fitted/predicted values $\widehat{y}$ of the dbh for each focal tree using the `predict_bayesian_model()`. These $\widehat{y}$ can then be compared to the observed $y$ dbh's to compute the root mean-square error, a measure of a model's predictive error which has the same units as the observed data $y$.

@@

```{r bw-model-fit}
comp_bayes_lm_bw <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL)
```

@@

```{r bw-model-predict}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  mutate(growth_hat = predict(comp_bayes_lm_bw, focal_vs_comp_bw))
```

@@

```{r bw-model-rmse}
focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r bw-model-shuffle-fit}
comp_bayes_lm_bw_shuffle <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)
```

```{r bw-model-shuffle-predict}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_bw_shuffle, focal_vs_comp_bw))
```

```{r bw-model-shuffle-rmse}
focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
```

@@

```{r scbi-model-fit}
comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
  comp_bayes_lm(prior_param = NULL)
```

```{r scbi-model-predict}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat = predict(comp_bayes_lm_scbi, focal_vs_comp_scbi))
```

```{r scbi-model-rmse}
focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-model-residuals, out.width="100%", fig.cap="Predicted versus observed growth.", echo = FALSE}
ggplot(focal_vs_comp_scbi, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() +
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh",
    title = "Predicted vs Observed Growth"
  )
```

```{r scbi-model-residuals-2, out.width="80%", fig.cap="Spatial distribution of residuals for model applied to SCBI data.", echo = FALSE}
focal_vs_comp_scbi %>%
  st_as_sf() %>%
  # TODO: Need to investigate missingness
  filter(!is.na(growth_hat)) %>%
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5),
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>%
  ggplot() +
  geom_sf(aes(col = error_compress), size = 1) +
  theme_bw() +
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf",
    name = expression(paste("Residual (cm ", y^{-1}, ")")),
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c("< -0.75", "-0.5", "0.25", "0", "0.25", "0.5", "> 0.75")
  ) +
  labs(x = "Meter", y = "Meter")
```

@@

```{r bw-posterior-viz-0}
focal_vs_comp_bw <- growth_bw %>%
  mutate(sp = family %>% factor()) %>%
  create_focal_vs_comp(comp_dist = comp_dist, cv_grid_sf = blocks_bw, id = "treeID")

@@

```{r bw-posterior-viz-1}
sp_to_plot <- c("cornaceae", "fagaceae", "hamamelidaceae", "juglandaceae", 
                 "lauraceae", "rosaceae", "sapindaceae", "ulmaceae")
```

@@

```{r bw-posterior-viz-beta0, out.width="100%", fig.cap="Posterior distribution of beta0."}
plot1 <- autoplot(comp_bayes_lm_bw, type = "intercepts")
plot1
```

@@

```{r bw-posterior-viz-beta-dbh, out.width="100%", fig.cap="Posterior distribution of betadbh."}
plot2 <- autoplot(comp_bayes_lm_bw, type = "dbh_slopes")
plot2
```

@@

```{r bw-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for Big Woods."}
plot3 <- autoplot(comp_bayes_lm_bw, type = "competition")
plot3
```

@@

```{r scbi-posterior-viz-1}
sp_to_plot <- c("quru", "litu", "cagl", "cato")
```

```{r scbi-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for SCBI."}
plot3 <- autoplot(comp_bayes_lm_bw, type = "competition")
plot3
```

@@

```{r bw-spatial-cv}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_bw)

@@

```{r scbi-spatial-cv}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)

>>>>

```{r setup, include = FALSE, cache = FALSE}
# Internally used packages
library(viridis)
library(knitr)
library(ggplot2)
library(here)
library(tibble)

# knitr settings
opts_chunk$set(
  # Code output:
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  cache = TRUE,
  # Figure:
  out.width = "100%",
  fig.path = "Figures/",
  fig.width = 16 / 2.5,
  fig.height = 9 / 2.5,
  fig.align = "center",
  fig.show = "hold",
  # Etc:
  collapse = TRUE,
  comment = "##",
  # tidy = FALSE,
  tibble.print_max = 5,
  tibble.print_min = 5
)

@@

## Model

Describe model specifics.

@@

1. Compute the growth of trees based on census data
1. Add spatial information:
    1. Define buffer region trees.
    1. Add spatial cross-validation block information.
1. Identify all focal and corresponding competitor trees.
1. Fit model and make predictions.
1. Additionally: Evaluate model performance using spatial cross-validation.
1. Additionally: Evaluate the effect of competitor species identity using permutation tests.

@@

The first step in the our analysis sequence is to compute the growth of trees using data from two censuses. The `compute_growth()` function computes growth assuming census data that follows ForestGEO standards. Despite such standards, minor variations will still exist between sites thereby necessitating some data wrangling and checking. For example, the SCBI site records all DBH's in millimeters, whereas the Michigan Big Woods site records them in centimeters @andersonteixeira_ctfs-forestgeo_2015 @allen_michigan_2020. The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all trees is saved in a data frame of class `sf` whereby the geolocation of each tree is presented in a `geometry` variable of type `<POINT>` and at a minimum the data contains the following variables: a variable uniquely identifying each tree-stem, `sp` of type `fct` factor identifying species, `dbh1` and `dbh2` of type `dbl` quantifying the DBH at earlier and later census, and `growth` of type `dbl` double quantifying the average annual growth in centimeters.

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020. After selecting only relevant variables, we perform a few additional data wrangling steps: convert the variable with the date of measurement to be of type `date`, convert DBH to be in centimeters^[A rule of thumb to determine the units of DBH is check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with DBH greater or equal to 1cm should be included in censuses.], convert the `sp` variable containing species information from type `chr` character to `fct` factor (we will discuss the need for this in Section \ref{spatial-cross-validation}). Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

The output `growth_scbi` is a single data frame of class `sf` that includes a numerical `growth` reflecting the average annual growth in DBH (in cm) for all trees that were alive at both time points as well a `geometry` variable encoding each tree's geolocation. Furthermore, variables that (in theory) remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`.

@@

## Identify all focal and corresponding competitor trees

The next step in our analysis sequence is to identify all focal trees and their corresponding competitor trees. The `create_focal_vs_comp()` functions performs these tasks and returns a new data frame of type `sf` containing this information. On top of the previously discussed arguments `comp_dist` defining the competition neighborhood and `id` indicating which variable in the data frame uniquely identifies each tree-stem, this function also requires an `sf` object representation of the spatial cross-validation blocks/folds; in our case, this was manually encoded in the `blocks_scbi` in Section \ref{spatial-information} while in our Appendix we present an example where this was performed using `spatialBlock()` from the `blockCV` package. We present the resulting data frame below with the `foldID` variable omitted for compactness of presentation.

```{r scbi-focal-vs-comp}
focal_vs_comp_scbi <- growth_scbi %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = blocks_scbi, id = "stemID")
focal_vs_comp_scbi %>% 
  select(-foldID)
```

TODO: Below reconcile the number of rows as they off by one from `growth_scbi` pipe `filter(!is.na(growth) & !buffer)`. perhaps by removing NA's in the `growth_scbi` stage.

The resulting data frame `focal_vs_comp_scbi` has `r nrow(focal_vs_comp_scbi)` rows, representing the subset of the `r nrow(growth_scbi)` trees in `growth_scbi` that will be considered as focal trees and thus have their growths modeled. Recall from Section \ref{spatial-information} this consists all trees that are not part of the buffer region in Figure \ref{fig:scbi-spatial-information}. Two new variables `focal_ID` and `focal_sp` related to tree-stem identification and species information. Most notably however is a new variable `comp` which contains information on all competitor trees for a given focal tree saved in list-column format, a feature of the `tidyr` package @tidyr_package. For example, we drill-down on the tree with `focal_ID` equal to 4. It has 20 competitor trees each described by 4 variables as indicated by the fact that `comp` is a `<tibble [20 × 4]>`.

```{r scbi-focal-vs-comp-2}
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp)
```

Using the `unnest()` function from the `tidyr` package, we can flatten list-column into regular columns. We observe that for the same focal tree with DBH equal to 13.65cm, we have information on all 20 competitor trees whose `dist` distance to the focal tree is less than or equal to `r comp_dist`, including their unique tree-stem ID number, their species, and their basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ where $DBH$ is the value from the earlier census in cm. Saving our focal versus competitor information in list-column minimizes redundancy since we do not repeat information on the focal tree 20 times.

```{r, eval = FALSE}
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp) %>% 
  unnest(cols = "comp")
```
```{r scbi-focal-vs-comp-3, echo = FALSE}
# Not sure why we need to do things this way:
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp) %>% 
  unnest(cols = "comp") %>% 
  print(n = 5)
```

@@

Next we fit the following linear model to the DBH of each focal tree. Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ "competitor" species groups. We modeled the growth in diameter per year $y_{ij}$ (in centimeters per year) of the $i^{th}$ tree of focal species group $j$ as a linear model $f$ of the following covariates $\vec{x}_{ij}$

@@

Using these posterior means, we then use the posterior predictive distribution to obtain fitted/predicted values $\widehat{y}$ of the DBH for each focal tree using the `predict_bayesian_model()`. These $\widehat{y}$ can then be compared to the observed $y$ DBH's to compute the root mean-square error, a measure of a model's predictive error which has the same units as the observed data $y$.

@@

```{r bw-model-fit, eval=FALSE}
comp_bayes_lm_bw <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL)
```

@@

```{r bw-model-predict, eval=FALSE}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  mutate(growth_hat = predict(comp_bayes_lm_bw, focal_vs_comp_bw))
```

@@

```{r bw-model-rmse, eval=FALSE}
focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r bw-model-shuffle-fit, eval=FALSE}
comp_bayes_lm_bw_shuffle <- focal_vs_comp_bw %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)
```

```{r bw-model-shuffle-predict, eval=FALSE}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_bw_shuffle, focal_vs_comp_bw))
```

```{r bw-model-shuffle-rmse, eval=FALSE}
focal_vs_comp_bw %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
```

@@

```{r scbi-model-fit, eval=FALSE}
comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
  comp_bayes_lm(prior_param = NULL)
```

```{r scbi-model-predict, eval=FALSE}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat = predict(comp_bayes_lm_scbi, focal_vs_comp_scbi))
```

```{r scbi-model-rmse, eval=FALSE}
focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

```{r scbi-model-residuals, out.width="100%", fig.cap="Predicted versus observed growth.", echo = FALSE, eval=FALSE}
ggplot(focal_vs_comp_scbi, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() +
  labs(
    x = "Observed growth in DBH", y = "Predicted growth in DBH",
    title = "Predicted vs Observed Growth"
  )
```

```{r scbi-model-residuals-2, out.width="80%", fig.cap="Spatial distribution of residuals for model applied to SCBI data.", echo = FALSE, eval=FALSE}
focal_vs_comp_scbi %>%
  st_as_sf() %>%
  # TODO: Need to investigate missingness
  filter(!is.na(growth_hat)) %>%
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5),
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>%
  ggplot() +
  geom_sf(aes(col = error_compress), size = 1) +
  theme_bw() +
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf",
    name = expression(paste("Residual (cm ", y^{-1}, ")")),
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c("< -0.75", "-0.5", "0.25", "0", "0.25", "0.5", "> 0.75")
  ) +
  labs(x = "Meter", y = "Meter")
```

@@

```{r bw-posterior-viz-0, eval=FALSE}
focal_vs_comp_bw <- growth_bw %>%
  mutate(sp = family %>% factor()) %>%
  create_focal_vs_comp(comp_dist = comp_dist, cv_grid_sf = blocks_bw, id = "treeID")

@@

```{r bw-posterior-viz-1, eval=FALSE}
sp_to_plot <- c("cornaceae", "fagaceae", "hamamelidaceae", "juglandaceae", 
                 "lauraceae", "rosaceae", "sapindaceae", "ulmaceae")
```

@@

```{r bw-posterior-viz-beta0, out.width="100%", fig.cap="Posterior distribution of beta0.", eval=FALSE}
plot1 <- autoplot(comp_bayes_lm_bw, type = "intercepts")
plot1
```

@@

```{r bw-posterior-viz-beta-dbh, out.width="100%", fig.cap="Posterior distribution of betadbh.", eval=FALSE}
plot2 <- autoplot(comp_bayes_lm_bw, type = "dbh_slopes")
plot2
```

@@

```{r bw-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for Big Woods.", eval=FALSE}
plot3 <- autoplot(comp_bayes_lm_bw, type = "competition")
plot3
```

@@

```{r scbi-posterior-viz-1, eval=FALSE}
sp_to_plot <- c("quru", "litu", "cagl", "cato")
```

```{r scbi-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for SCBI.", eval=FALSE}
plot3 <- autoplot(comp_bayes_lm_bw, type = "competition")
plot3
```

@@

```{r bw-spatial-cv, eval=FALSE}
focal_vs_comp_bw <- focal_vs_comp_bw %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_bw)

@@

```{r scbi-spatial-cv, eval=FALSE}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)
